---
title: "Workshop 8 TEMPLATE"
subtitle: "9 AM"
author: "Steven Le"
date: "2026-05-22"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest"
)

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Eschscholzia californica" = esccal_col,
  "Grindelia camporum" = gricam_col,
  "Stipa pulchra" = stipul_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Encelia californica" = enccal_col
)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating a new object
drought_exp_clean <- drought_exp |> 
  # cleaning names in object
  clean_names() |> 
  # creating new column to store binomial species names 
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # creating new column with full name for treatment 
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  )) |> 
  # reordering species levels 
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  # reordering water treatment levels
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting for desired columns
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: checking-structure

str(drought_exp_clean)
```

Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: ten-rows

slice_sample(
  drought_exp_clean,
  n = 10
)
```

# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 

```{r}

ggplot(data = drought_exp_clean,
       # SLA on x-axis and total mass on y-axis
       mapping = aes(x = sla,
                     y = total_g,
                     )) +
  # first layer: points
  geom_point() +
  # relabelling x- and y-axes
  # adding title
  labs(x = expression(Specific~Leaf~Area (cm^2/g)),
       y = "Total Mass (g)",
       title = "As SLA increases, total mass increases")

```

## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}

# base layer: ggplot with the x- and y-axes
ggplot(data = drought_exp_clean, # using the drought_exp_clean data frame
       aes(x = water_treatment, # x-axis
           y = total_g, # y-axis
           color = water_treatment)) + # coloring points by water_treatment
   # boxplot
   geom_boxplot() + 
  # first layer: showing the underlying data
  geom_jitter(height = 0, # no jitter in the vertical direction
              width = 0.1, # smaller jitter in the horizontal direction
              alpha = 0.4, # make the points more transparent
              shape = 21) + # make the points open circles
  # second layer: showing means with 95% CI
  labs(x = "Water Treatment", # labeling the axes
       y = "Total Mass (g)",
       title = "Median total mass (g) for well watered treatments is higher than drought\nstressed treatments") +
  # figure customization to change the point colors
  # Note: this is optional (but nice to do!)
  scale_color_manual(values = c("Well watered" = "darkorchid3",
                                "Drought stressed" = "tomato3")) + 
  theme_bw() + # using a theme
  theme(legend.position = "none") # getting rid of the legend

```

## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.

```{r}

# base layer: ggplot with the x- and y-axes
ggplot(data = drought_exp_clean, # using the drought_exp_clean data frame
       aes(x = species_name, # x-axis
           y = total_g, # y-axis
           color = species_name)) + # coloring points by species_name
   # boxplot
   geom_boxplot() + 
  # first layer: showing the underlying data
  geom_jitter(height = 0, # no jitter in the vertical direction
              width = 0.1, # smaller jitter in the horizontal direction
              alpha = 0.4, # make the points more transparent
              shape = 21) + # make the points open circles
  # second layer: showing means with 95% CI
  labs(x = "Species", # labeling the axes
       y = "Total Mass (g)",
       title = "Median total plant mass (g) varies between species") +
  scale_color_manual(values = plant_colors) + 
  theme_bw() + # using a theme
  theme(legend.position = "none") # getting rid of the legend

```

# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(
  total_g ~ 1,
  data = drought_exp_clean
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name,
  data = drought_exp_clean
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment,
  data = drought_exp_clean
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean
)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ water_treatment + species_name,
  data = drought_exp_clean
)

# model 5: SLA
model5 <- lm(
  total_g ~ sla,
  data = drought_exp_clean
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)

# model 7: species
model7 <- lm(
  total_g ~ species_name,
  data = drought_exp_clean
)
```

## Model diagnostics

```{r}
#| label: check-diagnostics 

# displaying plots in 2x2 grid
par(mfrow = c(2, 2))
plot(model4)
```

- residuals appear normally distributed
- residuals appear homoscedastic (have constant variance)
- no clear outliers in data set influencing model coefficient estimates 

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: calculating-aic

AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  #arranging output in order of AIC
  arrange(AICc)
```

Insert code to display the summary for the best model. 

```{r}

summary(model4)
```

- '(Intercept)': model prediction for plant mass for Lotus scoparius in drought stressed treatment 
- 'water_treatmentWell watered': model prediction for difference in Lotus scoparius mass between well watered and drought stressed treatment.
- 'species_namePenstemon centranthifolius': model prediction for difference between Penstemon mass in well watered treatment and lotus mass in well watered treatment 

## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors.

**How much variation in the response (total mass, in grams) does this model explain?**

72% of the variation in plant mass is explained by a model with both of these predictors.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

- Treatment influences plant mass.
- Species infleunces plant mass.

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: model-predictions


preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |> 
  rename(water_treatment = x,
         species_name = group)
```

# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
#| label: final-figure

ggplot(data = drought_exp_clean,
       mapping = aes(x = water_treatment, 
                     y = total_g,
                     color = water_treatment)) +
  geom_jitter(height = 0,
              width = 0.1) + 
  geom_pointrange(data = preds,
                  mapping = aes(x = water_treatment,
                                y = predicted,
                                ymin = conf.low,
                                ymax = conf.high)) +
  facet_wrap(~ species_name) +
  theme(legend.position = "none")
```

# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







